Type theory in type theory with single substitutions

### 2024-09-15 types_2024_postproc/p.tex
minimisation

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:ssc}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

@@

TODO. In Section \ref{sec:ssc}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

\section{Single substitution syntax}
\label{sec:ssc}

@@

So far our only terms are variables, but we would like to define
functions via lambda abstraction. Abstraction takes a term in a
context extended by the domain of the function. It comes with a rule
for instantiation analogous to $\Pi[]$.
\begin{alignat*}{10}
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+])
\end{alignat*}
Note that the equation $\lam[]$ only makes sense because of the
previous equation $\Pi[]$: the left hand side is in
$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
right hand side is in
$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the typal version of the
equation mentioned in the introduction.
\begin{alignat*}{10}
& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

@@

@@

In this section we show that a category with families (CwF) equipped
with a hierarchy of Coquand-universes and $\Pi$ types can be derived
from the single substitution syntax defined in \autoref{sec:ssc}.

@@

In this section we show that if have we universes and $\Pi$ types
(which is the case in the theory (\ref{eq:tt})), we can decrease the
number of euqations.

In our single substitution calculus, we have the equation $[\p][^+]$
stated both for types and terms (we know that the left one is about
types because of our convention that type metavariables are uppercase
Latin letters and term metavariables are lowercase):
\[
B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
\]
We even need the first equation to typecheck the second one. We made
the dependency explicit by adding a subscript of the equality in the
equation for terms (in Agda, this dependency has to be made
explicit). An alternative presentation of the second equation without
requiring the first one is the following:
\[
[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
\]
That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
we have that for any $b$ we have the equation (suitably over the input
equation). Thus we add a conditional equation.

It turns out that if we have Coquand-universes, the conditional
$[\p][^+]'$ rule is enough: the input equation holds for $B := \U\,i$
via $\U[]$, thus we get that for any $\hat{A} : \Tm\,\Gamma\,(\U\,i)$,
$\hat{A}[\p][\gamma^+] = \hat{A}[\gamma][\p]$. But every type has a
code in the universe, so for any $B : \Ty\,\Gamma\,i$,
\begin{alignat*}{10}
  & B[\p][\gamma^+] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]') \\
  & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
  & \El\,(\cd\,B)[\gamma][\p] && {=}(\U\beta) \\
  & B[\gamma][\p].
\end{alignat*}
Equation $[\q][^+] : \q[\gamma^+] = \q$ only makes sense if we have
$[\p][^+]$ for types, so either we make $[\q][^+]$ conditional, or we
coerce it along the derived equation.

We play the exact same game with $[\p][\langle\rangle]$: we replace it
with the conditional equation
\[
[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b,
\]
notice that we have the input for $B = \U\,i$, and derive the input
equation for all types.

Now we turn our attention to the equation
\[
  [\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle].
\]
We were forced to introduce it so that we can state the substitution
rule for function application $\cdot[]$. What if we made $\cdot[]$
itself conditional?

xxx

@@

@@

\section{The single substitution GAT presentation of the SOGAT (\ref{eq:tt})}

For reference, we list the rules of the theory introduced in Section
\ref{sec:ssc}. The core substitution calculus:
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
& \Sub && : \Con\ra\Con\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
& \q[^+] && : \q[\gamma^+] = \q \\
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
\end{alignat*}
Rules for individual type formers.
\begin{alignat*}{10}
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}

>>>>

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

@@

TODO. In Section \ref{sec:tt}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

\section{Single substitution syntax}
\label{sec:tt}

@@

So far our only terms are variables, but we would like to define
functions via lambda abstraction. Abstraction takes a term in a
context extended by the domain of the function. It comes with a rule
for instantiation analogous to $\Pi[]$.
\begin{alignat*}{10}
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
\end{alignat*}
Note that the equation $\lam[]$ only makes sense because of the
previous equation $\Pi[]$: the left hand side is in
$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
right hand side is in
$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the typal version of the
equation mentioned in the introduction.
\begin{alignat*}{10}
& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

@@

TODO: beta equation for id.

@@

In this section we show that a category with families (CwF) equipped
with a hierarchy of Coquand-universes and $\Pi$ types can be derived
from the single substitution syntax defined in \autoref{sec:tt}.

@@

In this section we show that if have we universes and $\Pi$ types
(which is the case in the theory (\ref{eq:tt})), we can decrease the
number of equations.

In our single substitution calculus, we have equation $[\p][^+]$
stated both for types and terms (we know that the left one is about
types because of our convention that type metavariables are uppercase
Latin letters and term metavariables are lowercase):
\[
B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
\]
We even need the first equation to typecheck the second one. We made
the dependency explicit by adding a subscript of the equality in the
equation for terms (in Agda, this dependency has to be made
explicit). An alternative presentation of the second equation without
requiring the first one is the following:
\[
[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
\]
That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
we have that for any $b$ we have the equation (suitably over the input
equation). Thus we add a conditional equation.

It turns out that if we have Coquand-universes, the conditional
$[\p][^+]'$ rule is enough: the input equation holds for $B := \U\,i$
via $\U[]$, thus we get that for any $\hat{A} : \Tm\,\Gamma\,(\U\,i)$,
$\hat{A}[\p][\gamma^+] = \hat{A}[\gamma][\p]$. But every type has a
code in the universe, so for a $B : \Ty\,\Gamma\,i$,
\begin{alignat*}{10}
  & B[\p][\gamma^+] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]'\,\U[]) \\
  & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
  & \El\,(\cd\,B)[\gamma][\p] && {=}(\U\beta) \\
  & B[\gamma][\p].
\end{alignat*}
Equation $[\q][^+] : \q[\gamma^+] = \q$ also only makes sense if we
have $[\p][^+]$ for types, so either we have to make $[\q][^+]$
conditional, or we coerce it along the derived equation.

We play the exact same game with $[\p][\langle\rangle]$: we replace it
with the conditional equation
\[
[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b,
\]
notice that we have the input for $B = \U\,i$, and derive the input
equation for all types. We make $\q[\langle\rangle]$ conditional as
well.

Now we turn our attention to the equation
\[
  [\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle].
\]
We were forced to introduce it so that we can state the substitution
rule for function application $\cdot[]$. What if we made $\cdot[]$
itself conditional?
\begin{alignat*}{10}
  & \cdot[]' :{} && (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
  & && (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma])
\end{alignat*}
Again, for $B = \U$ we have the assumption $e$. But then $t$ is in
$\Tm\,\Gamma\,(\Pi\,A\,\U)$, which is isomorphic to $\Ty\,(\Gamma\ext
A)$. So we argue as follows for any $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\langle a\rangle][\gamma] && {=}(\U\beta) \\
  & \El\,(\cd\,(B[\langle a\rangle][\gamma])) && {=}(\cd[]) \\
  & \El\,((\cd\,B)[\langle a\rangle][\gamma]) && {=}(\Pi\beta) \\
  & \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) && {=}(\cdot[]'\,\U[]) \\
  & \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma]))\,\, && {=}(\lam[]) \\
  & \El\,(\lam\,(\cd\,B[\gamma^+])\cdot (a[\gamma])) && {=}(\Pi\beta) \\
  & \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) && {=}(\cd[]) \\
  & \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] && {=}(\U\beta) \\
  & B[\gamma^+][\langle a[\gamma]\rangle]
\end{alignat*}
Hence, the conditional $\cdot[]'$ equation implies the condition for
all types.

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we exchange
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
With this we derived an even more minimalistic first-order theory for
the SOGAT (\ref{eq:tt}), essentially relying on the universe and $\Pi$
types. It is clear from the construction that the resulting theory is
isomorphic to the one presented in Section \ref{sec:tt} (and Appendix
\ref{app:tt}), all equations are interderivable. For reference, we
list all the rules of the minimised calculus in Appendix
\ref{app:minimisation}.

@@

intuition, similar to minimisation: the subst calc is just there to
bootstrap U and Pi, those can do everything.

@@

\section{GAT presentation of (\ref{eq:tt})}
\label{app:tt}

For reference, we list the rules of the theory introduced in Section
\ref{sec:tt}. The core substitution calculus:
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
  & [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
  & \q[^+] && : \q[\gamma^+] = \q \\
  & [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
  & [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
  & \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
  & [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
  & [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
\end{alignat*}
Rules for individual type formers.
\begin{alignat*}{10}
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])
\end{alignat*}
\begin{alignat*}{10}
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}

\section{Minimised GAT presentation of (\ref{eq:tt})}
\label{app:minimisation}

We list the rules of the theory derived in Section
\ref{sec:minimisation}.
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_e b[\gamma][\p] \\
  & \q[^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra \q[\gamma^+] =_e \q \\
  & [\p][\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra b[\p ][\langle a\rangle] =_e b \\
  & \q[\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra \q[\langle a\rangle] =_e a \\
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[]' && : (e:B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
  & && \hphantom{{}:{}} (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
